---
title: "Exploratory Analysis"
author: "Abdullahi Adinoyi IBRAHIM \n Jean-Paul ISHIMWE "
date: "19 September 2018"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

# Exploratory Analysis

## Loading the data
```{r}
data <- read.csv("C:\\Users\\abdon\\Documents\\R\\BRD projects guide\\internship\\Assignmet\\Copy of laterite_education_data.csv", sep = ",", header=TRUE )

data$Column1 <- NULL 
```

## Question 1             
Renaming the features
```{r}
colnames(data) <- c("province", "district", "urban_rural_class", "weight", "consumption",
                    "sex", "Age", "father_alive", "mother_alive", "healthproblem_4weeks", 
                    "gradeAttended_2012", "gradeAttended_2013", "prevSch_12months", 
                    "problem_experienced", "total_Education_Expenses", "WhoPaidExpenses_prev_12months",
                    "num_missedSch_7days", "Reason4notattend", "Reason4leaving", "Can_Read", "Can_write",
                    "can_Perform_calculation", "work_on_household_farm" )
```

             
## Question 2                   
                     
Creating the dummy variables

```{r}
primary_only <- data[data$gradeAttended_2012=="Primary 1" | data$gradeAttended_2012=="Primary 2" | data$gradeAttended_2012=="Primary 3" | data$gradeAttended_2012=="Primary 4" | data$gradeAttended_2012=="Primary 5" | data$gradeAttended_2012=="Primary 6", ]

# repeated
primary_only$repeated <- ifelse(primary_only$gradeAttended_2012 == "Primary 1" & primary_only$gradeAttended_2013=="Primary 1",                                  "repeated",
                         ifelse(primary_only$gradeAttended_2012 == "Primary 2" & primary_only$gradeAttended_2013=="Primary 2",                                 "repeated",
                         ifelse(primary_only$gradeAttended_2012 == "Primary 3" & primary_only$gradeAttended_2013=="Primary 3",                                "repeated",
                         ifelse(primary_only$gradeAttended_2012 == "Primary 4" & primary_only$gradeAttended_2013=="Primary 4",                                "repeated",
                         ifelse(primary_only$gradeAttended_2012 == "Primary 5" & primary_only$gradeAttended_2013=="Primary 5",                                "repeated",
                         ifelse(primary_only$gradeAttended_2012 == "Primary 6" & primary_only$gradeAttended_2013=="Primary 6",                                "repeated", "Promoted" )))))) 



# dropout
primary_only$dropout <- ifelse(primary_only$gradeAttended_2012 =="Primary 1" & primary_only$gradeAttended_2013=="Not in class",                              "dropout",
                      ifelse(primary_only$gradeAttended_2012 =="Primary 2" & primary_only$gradeAttended_2013=="Not in class",                                "dropout",
                      ifelse(primary_only$gradeAttended_2012 =="Primary 3" & primary_only$gradeAttended_2013=="Not in class",                               "dropout",
                      ifelse(primary_only$gradeAttended_2012 =="Primary 4" & primary_only$gradeAttended_2013=="Not in class",                               "dropout",
                      ifelse(primary_only$gradeAttended_2012 =="Primary 5" & primary_only$gradeAttended_2013=="Not in class",                               "dropout",
                       ifelse(primary_only$gradeAttended_2012 =="Primary 6" & primary_only$gradeAttended_2013=="Not in class",                              "dropout","In School"))))))

# table for dropout
library(gmodels)
CrossTable(primary_only$dropout)

# table for repeared
CrossTable(primary_only$repeated)
```
                                 
* the overall dropout rate for dropout is $5.2\%$ and grade repetition is $25.1\%$  
                    
                     
##  Question 3
```{r}
library(gmodels)
library(ggplot2)

# table for repeated
CrossTable(primary_only$gradeAttended_2012, primary_only$repeated,prop.r = TRUE, prop.c = FALSE, prop.t = FALSE, prop.chisq = TRUE)

# Visaulization for population
ggplot(primary_only) + 
  geom_bar(aes(x=gradeAttended_2012),fill = 'gray') + 
  coord_flip() +
  theme(axis.text.y = element_text(size = rel(.8))) +ggtitle("Population of student")

# Visaulization for repeated
ggplot(primary_only) + 
  geom_bar(aes(x=repeated),fill = 'gray') + 
  coord_flip() +
  theme(axis.text.y = element_text(size = rel(.8))) +ggtitle("Repeated and not repeated")
```
                         
* From our analysis, the grade repetition in primary education is sumaarised as follow; In primary 1 $39.5\%$ repeated, $23.6\%$ repeated in Primary 2, $16.8\%$ repated in primary 3, $16.7\%$ in Primry 4, $22.3\%$ in Primary 5 and $6.8\%$ repeated in Primary 6. Overall number of students that repeated is 567 out of 2260 students for the year 2012.                           

                   
## Question 4
```{r}
attach(primary_only)

# Visualization of dropout
ggplot(primary_only) + 
  geom_bar(aes(x=dropout,fill = sex)) + 
  coord_flip() +
  theme(axis.text.y = element_text(size = rel(.8))) +ggtitle("Repeated and not repeated")

# Table
CrossTable(primary_only$sex, primary_only$dropout,prop.r = TRUE, prop.c = FALSE, prop.t = FALSE, prop.chisq = FALSE)
```
                 
* From analysis above, the dropout rate is $5\%$ for Female and $5.4\%$ for Male, the difference between these dropout rate is $0.4\%$ which is statistically insignificant, using the knowledge of confidence interval, the difference between the error rates is $0.4\%$ < $5% CI%.                     

* This means that the dropout rate between male and female students is almost thesame since we can round down $5.4$ to $5$, hence we can say that the dropout rate between male and female is aprroximately the same rate.


## Question 5
```{r}
repeated_students <- primary_only[primary_only$repeated == "repeated",]
#summary(repeated_students)
```

# Another technique
```{r}
library(party)
library(randomForest)
library(caret)
# randomforest
primary_only$repeated <- as.factor(primary_only$repeated)
primary_only$dropout <- as.factor(primary_only$dropout)

rf <- cforest(repeated ~., data=primary_only, control=cforest_unbiased(mtry=2,ntree=50))
rf1 <- randomForest(repeated ~., data = primary_only)
```

```{r}
set.seed(1)
options(scipen = 999)
# Important features
varimp(rf)
#importance(rf)
varImp(rf)
```

                   
* The following are good predictors of repetition in primary Education.
       * Can_read
       * Can_write
       * can_perform_calculation
       * Number of school missed in last 7 days
       *  healthproblem_4week
       * Reason4notattend
  
* Out of `r nrow(repeated_students)`, $235$ students **can not read** and 128 students in the "Can_read" column is not classified, this means out of the unclassided students some will fall in the can not read class. Overall number of students that can not read is $ > 235$ students

* Out of `r nrow(repeated_students)`, $220$ pupils **can not perform calculation** and $128$ pupils are not classified also, the overall number of students that can not perform calculation will be more than $220$ if the unclassified pupils are classified
           
* for "no_missedSch_7days" column, many students did not attend class in the last 7 days and also $336$ students are not classified. this feature is another major factor for repetition since not attending class is a sign of unseriousness in the student   
                   
* $16\%$ of students have health issues in the last 4 weeks





## Question 6                
```{r}
dropout_students <- primary_only[primary_only$dropout == "dropout",]
summary(dropout_students)
```
                  
### Strengths              
               
* this dataset covers all the provinces in Rwanda.                              
                                                                                                   
* The data has broad scope, this is because the data is given according to the districts in Rwanda                 
                           
* The "gradeAttended_2012" feature is rich, allowing us to classify students that dropout of class according to their classes


### Limitations                 
               
* The data is missing the total earnings of the sponsor of the student, we need this information to determine the cause of dropout. 
       
* In the "problem_expericed" feature, information to the kind of problem experienced by the student is not given, this would have assist to determine what kind of problem made students dropout of school
            
* "WhoPaidExpenses_12months" feature is also not given, this is important in other to determine why a student dropout, it could be due to incapacity of the sponsor that made the student dropout of school
             
* The "Total Education expenses" feature is missing lots of information, if the total expenditure of student are known, we could determine maybe its due to high expenses that made a student 



